Windowed least-squares model reduction for dynamical systems

### 2020-10-09 numerical_experiments.tex
latest updates

\subsubsection{Description of FOM and generation of \spatialAcronym\ trial subspace}\label{sec:sod_fom}
We solve the 1D compressible Euler equations with a finite volume method. We partition the domain into 500 cells of uniform width and employ the Rusanov flux~\cite{rusanov} at the cell interfaces. We employ the Crank--Nicolson (CN) scheme, which is a linear multistep method defined by the coefficients $\alpha_0 = 1,\alpha_1 = -1, \beta_0 = \beta_1 = 1/2$, for temporal integration. We evolve the FOM for $t \in [0.0,1.0]$ at a time-step of $\Delta t = 0.002$. We construct the \spatialAcronym\ trial subspace by executing Algorithm~\ref{alg:pod} with inputs $N_{\text{skip}}=2, \stateIntercept = \bz, K=46 $. The resulting trial subspace corresponds to an energy criterion of $99.99\%$.
%  The solution is saved every other time-step, resulting in 500 solutions snapshots. These snapshots are used to generate a trial basis for the reduced-order models through POD. The trial basis is selected to be of dimension $K = 46$, which corresponds to a $99.99\%$ energy criterion. No affine transformation is used in the construction of the trial subspace, i.e., $\stateIntercept = \bz$.

@@

\subsubsection{Numerical results}
We first assess the impact of the window size on the performance of the \methodAcronymROMs.  We consider a set of \methodAcronymROMs\ that minimize the residual over windows of constant size 
$\DeltaSlabArg{n} \equiv \DeltaSlabArg{} =  .002$, $0.004$ ,$0.008$, $0.02$, $0.04$, $0.10$, $0.20$, $1.0$. We additionally consider the standard Galerkin and LSPG ROMs. We first show results for \methodAcronymROMs\ using the direct method with CN time discretization; a comparison of different time-marching methods and direct/indirect solution techniques will be provided later in this section. 
First, Figure~\ref{fig:sod_density} presents the density solutions produced by the various ROMs at $t = 0.5$ and $1.0$. Figure~\ref{fig:sod_xt} shows 
$\mathsf{x}-t$ diagrams for the same density solutions. From Figures~\ref{fig:sod_density} and~\ref{fig:sod_xt}, we observe that the LSPG and Galerkin ROMs accurately characterize 
the system: they correctly track the shock location, expansion waves, etc. We observe both predictions, however, to be highly oscillatory. These oscillations are 
not physical and can lead to numerical instabilities; e.g., due to negative pressure. We observe the \methodAcronymROMs\ to produce less oscillatory solutions than both 
the Galerkin and LSPG ROMs. Critically, we see that the solution becomes less oscillatory as the window size over which the residual is minimized grows. The solution displays no oscillations when the residual is minimized over the entire space--time domain.

@@

Next, Figure~\ref{fig:sod_error} shows space--time state errors and the objective function~\eqref{eq:obj} (i.e., the space--time residual norm) over $t \in [0,1]$ for the various ROMs. Most notably, we observe 
that increasing the window size over which the residual is minimized does \textit{not} lead to a monotonic decrease in the space--time error as measured in the $\elltwo$-norm (we refer to this as the $\elltwo$-error). As expected, increasing the window size does lead to a monotonic decrease in the space--time residual, however. We additionally note that, although the space--time
$\elltwo$-error of 
the projected FOM solution is significantly lower than that of the various ROM solutions, the space--time residual norm of the projected FOM solution is \textit{higher} 
than all ROMs. Thus, although the projected FOM solution is more accurate in the $\elltwo$-norm, it does not satisfy the governing equations as well the ROM solutions.

\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/error_vs_window_compare.pdf}
\caption{Space--time $\elltwo$-error of various ROMs.}
\label{fig:sod_error_a}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/objective_vs_window_compare.pdf}
\caption{Integrated residual norm of various ROMs.} 
\label{fig:sod_error_b}
\end{subfigure}
\caption{Integrated performance metrics of the Galerkin, LSPG, and \methodAcronymROMs.  Note that the Galerkin and LSPG ROMs do not depend on the window size.} 
\label{fig:sod_error}
\end{center}
\end{figure}

We now examine the comparative performance of the direct and indirect solution techniques for the \methodAcronymROMs\ using various time discretization techniques. 
Figure~\ref{fig:sod_error_methods} 
shows the same space--time $\elltwo$-errors and residual norms as in Figure~\ref{fig:sod_error}, but this time results are shown for 
the various \methodAcronymROMs. In both Figures~\ref{fig:sod_error_methods_a} and~\ref{fig:sod_error_methods_b}, we observe that the \methodAcronym\ method 
is relatively insensitive to the solution technique (direct vs indirect) and underlying discretization scheme, although some minor differences are observed.
In particular, ROMs using the second order explicit AB2 scheme with a time step of $\Delta t = 0.0005$ provide similar results to the 
ROMs using the second-order CN scheme at a time step of $\Delta t = 0.002$. All methods display similar dependence on the window size: the optimal 
$\elltwo$-error occurs when the window size is $\DeltaSlabArg{} = 0.1$, and the residual decreases monotonically as the window size grows. 
\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/error_vs_window.pdf}
\caption{Integrated $\elltwo$-error of the \methodAcronymROMs.}
\label{fig:sod_error_methods_a}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/objective_vs_window.pdf}
\caption{Integrated residual norm of the \methodAcronymROMs.} 
\label{fig:sod_error_methods_b}
\end{subfigure}
\caption{Integrated performance metrics of various \methodAcronymROMs.} 
\label{fig:sod_error_methods}
\end{center}
\end{figure}

@@

Next, we study the impact of the time step on the \methodAcronymROM\ results. We examine \methodAcronymROMs\ that use a
window size of $\Delta T = 0.1$, with time steps $\Delta t =$  $0.001$, $0.002$, $0.005$, $0.01$, $0.05$, $0.1$ (i.e.,  
$100$, $50$, $20$, $10$, $2$, and $1$ time instances per window). We additionally 
consider LSPG ROMs leveraging the same set of time steps. To assess time step convergence, we compare results to a new full-order model, which is as described in Section~\ref{sec:sod_fom} but uses a fine time step of $\Delta t = 10^{-4}$. Figure~\ref{fig:time_step_study} shows the $\elltwo$-error and residual norm  
of the various ROMs. We observe that the $\elltwo$-error and residual norm 
of the \methodAcronymROMs\ decrease and converge as the time step decreases. This is in direct contrast to the LSPG ROMs, in where the $\elltwo$-error and residual norm display a complex dependence on the time step. This result 
demonstrates that \methodAcronym\ overcomes the time-step dependence inherit to the LSPG approach. Lastly, Figure~\ref{fig:walltime_dtvar}
shows the relative wall-clock times of the \methodAcronymROMs\ with respect to the LSPG ROMs. It is seen that for all time steps considered, the \methodAcronymROMs\ 
are less than 6x the cost of LSPG; this is despite the window sizes comprising up to 100 time instances. 
\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/error_vs_window_dtvar.pdf}
\caption{Space--time $\elltwo$-error of the \methodAcronymROM.}
\label{fig:sod_error_methods_a}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/objective_vs_window_dtvar.pdf}
\caption{Space--time residual norm of the \methodAcronymROM.} 
\label{fig:sod_error_methods_b}
\end{subfigure}
\caption{Performance metrics of the Galerkin, LSPG, and \methodAcronymROMs.} 
\label{fig:time_step_study}
\end{center}
\end{figure}

@@

Figure~\ref{fig:convergence_study} shows the integrated $\ell^2$-error of the various ROM solutions, as well as the optimal solution as measured in the $\ell^2$-norm. In Figure~\ref{fig:sod_error_converge} we first observe that for a basis of dimension $\romdim = 10$, the Galerkin, LSPG, and WLS $\Delta T = 0.002$, $\Delta T = 0.004$ ROMs do not converge. Upon investigation, this lack of convergence occurred due to the presence of negative pressure. We next observe that the $\ell^2$-error monotonically decreases as the basis dimension increases for all remaining ROM solutions with the exception of the Galerkin, LSPG, and WLS $\Delta T = 0.002$, $\Delta T = 0.004$ at dimension $\romdim = 30$. Overall, we observe that larger window sizes yield lower errors when the basis dimension is small. For larger basis dimensions we observe all ROMs to yield accurate solutions, and a larger window size does not always yield a lower error. In Figure~\ref{fig:sod_residual_converge}, we next observe that the residual monotonically decreases as the basis dimension grows. This result is supported, although not guaranteed, by the error analysis presented in Section~\ref{sec:analysis} and Ref.~\cite{carlberg_lspg_v_galerkin}. Finally, we observe that larger window sizes yields lower integrated residuals.

\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/error_converge.pdf}
\caption{Space--time $\elltwo$-error of the \methodAcronymROM.}
\label{fig:sod_error_converge}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/residual_converge.pdf}
\caption{Space--time residual norm of the \methodAcronymROM.} 
\label{fig:sod_residual_converge}
\end{subfigure}
\caption{Performance metrics of the Galerkin, LSPG, and \methodAcronymROMs\ as a function of basis dimension. Missing entries indicate that the solver did not converge.} 
\label{fig:convergence_study}
\end{center}
\end{figure}

>>>>

\subsubsection{Description of FOM and generation of \spatialAcronym\ trial subspace}\label{sec:sod_fom}
We solve the 1D compressible Euler equations with a finite volume method. We partition the domain into 500 cells of uniform width and employ the Rusanov flux~\cite{rusanov} at the cell interfaces. We employ the Crank--Nicolson (CN) scheme, which is a linear multistep method defined by the coefficients $\alpha_0 = 1,\alpha_1 = -1, \beta_0 = \beta_1 = 1/2$, for temporal integration. We evolve the FOM for $t \in [0.0,1.0]$ at a time-step of $\Delta t = 0.002$. We construct the \spatialAcronym\ trial subspace by executing Algorithm~\ref{alg:pod} with inputs $N_{\text{skip}}=2, \stateIntercept = \bz, K=46 $, and $\mathbf{P} = \mathbf{I}$. The resulting trial subspace corresponds to an energy criterion of $99.99\%$.
%  The solution is saved every other time-step, resulting in 500 solutions snapshots. These snapshots are used to generate a trial basis for the reduced-order models through POD. The trial basis is selected to be of dimension $K = 46$, which corresponds to a $99.99\%$ energy criterion. No affine transformation is used in the construction of the trial subspace, i.e., $\stateIntercept = \bz$.

@@

\subsubsection{Numerical results}
We first assess the impact of the window size on the performance of the \methodAcronymROMs.  We consider a set of \methodAcronymROMs\ that minimize the residual over windows of constant size 
$\DeltaSlabArg{n} \equiv \DeltaSlabArg{} =  0.002$, $0.004$ ,$0.008$, $0.02$, $0.04$, $0.10$, $0.20$, $1.0$. We additionally consider the standard Galerkin and LSPG ROMs. We first show results for \methodAcronymROMs\ using the direct method with CN time discretization; a comparison of different time-marching methods and direct/indirect solution techniques will be provided later in this section. 
First, Figure~\ref{fig:sod_density} presents the density solutions produced by the various ROMs at $t = 0.5$ and $1.0$. Figure~\ref{fig:sod_xt} shows 
$\mathsf{x}-t$ diagrams for the same density solutions. From Figures~\ref{fig:sod_density} and~\ref{fig:sod_xt}, we observe that the LSPG and Galerkin ROMs accurately characterize 
the system: they correctly track the shock location, expansion waves, etc. We observe both predictions, however, to be highly oscillatory. These oscillations are 
not physical and can lead to numerical instabilities; e.g., due to negative pressure. We observe the \methodAcronymROMs\ to produce less oscillatory solutions than both 
the Galerkin and LSPG ROMs. Critically, we see that the solution becomes less oscillatory as the window size over which the residual is minimized grows. The solution displays no oscillations when the residual is minimized over the entire space--time domain.

@@

Next, Figure~\ref{fig:sod_error} shows time-integrated $\ell^2$-norm of the state errors and the objective function~\eqref{eq:obj} (i.e., the time-integrated residual norm) over $t \in [0,1]$ for the various ROMs. Most notably, we observe 
that increasing the window size over which the residual is minimized does \textit{not} lead to a monotonic decrease in the error as measured in the time-integrated $\elltwo$-norm (we refer to this as the $\elltwo$-error). As expected, increasing the window size does lead to a monotonic decrease in the time-integrated residual norm, however. We additionally note that, although the time-integrated $\elltwo$-error of 
the projected FOM solution is significantly lower than that of the various ROM solutions, the time-integrated residual norm of the projected FOM solution is \textit{higher} 
than all ROMs. Thus, although the projected FOM solution is more accurate in the $\elltwo$-norm, it does not satisfy the governing equations as well the ROM solutions.

\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/error_vs_window_compare.pdf}
\caption{Relative time-integrated $\elltwo$-error of various ROMs.}
\label{fig:sod_error_a}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/objective_vs_window_compare.pdf}
\caption{Time-integrated residual norm of various ROMs.} 
\label{fig:sod_error_b}
\end{subfigure}
\caption{Time-integrated performance metrics of the Galerkin, LSPG, and \methodAcronymROMs.  Note that the Galerkin and LSPG ROMs do not depend on the window size.} 
\label{fig:sod_error}
\end{center}
\end{figure}

We now examine the comparative performance of the direct and indirect solution techniques for the \methodAcronymROMs\ using various time discretization techniques. 
Figure~\ref{fig:sod_error_methods} 
shows the same time-integrated $\elltwo$-errors and residual norms as in Figure~\ref{fig:sod_error}, but this time results are shown for 
the various \methodAcronymROMs. In both Figures~\ref{fig:sod_error_methods_a} and~\ref{fig:sod_error_methods_b}, we observe that the \methodAcronym\ method 
is relatively insensitive to the solution technique (direct vs indirect) and underlying discretization scheme, although some minor differences are observed.
In particular, ROMs using the second order explicit AB2 scheme with a time step of $\Delta t = 0.0005$ provide similar results to the 
ROMs using the second-order CN scheme at a time step of $\Delta t = 0.002$. All methods display similar dependence on the window size: the optimal time-integrated 
$\elltwo$-error occurs when the window size is $\DeltaSlabArg{} = 0.1$, and the time-integrated residual norm decreases monotonically as the window size grows. 
\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/error_vs_window.pdf}
\caption{Relative time-integrated $\elltwo$-error of the \methodAcronymROMs.}
\label{fig:sod_error_methods_a}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/objective_vs_window.pdf}
\caption{Time-integrated residual norm of the \methodAcronymROMs.} 
\label{fig:sod_error_methods_b}
\end{subfigure}
\caption{Time-integrated performance metrics of various \methodAcronymROMs.} 
\label{fig:sod_error_methods}
\end{center}
\end{figure}

@@

Next, we study the impact of the time step on the \methodAcronymROM\ results. We examine \methodAcronymROMs\ that use a
window size of $\Delta T = 0.1$, with time steps $\Delta t =$  $0.001$, $0.002$, $0.005$, $0.01$, $0.05$, $0.1$ (i.e.,  
$100$, $50$, $20$, $10$, $2$, and $1$ time instances per window). We additionally 
consider LSPG ROMs leveraging the same set of time steps. To assess time step convergence, we compare results to a new full-order model, which is as described in Section~\ref{sec:sod_fom} but uses a fine time step of $\Delta t = 10^{-4}$. Figure~\ref{fig:time_step_study} shows the time-integrated $\elltwo$-error and residual norm  
of the various ROMs. We observe that the time-integrated $\elltwo$-error and residual norm 
of the \methodAcronymROMs\ decrease and converge as the time step decreases. This is in direct contrast to the LSPG ROMs, in where the time-integrated $\elltwo$-error and residual norm display a complex dependence on the time step. This result 
demonstrates that \methodAcronym\ overcomes the time-step dependence inherit to the LSPG approach. Lastly, Figure~\ref{fig:walltime_dtvar}
shows the relative wall-clock times of the \methodAcronymROMs\ with respect to the LSPG ROMs. It is seen that for all time steps considered, the \methodAcronymROMs\ 
are less than 6x the cost of LSPG; this is despite the window sizes comprising up to 100 time instances. 
\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/error_vs_window_dtvar.pdf}
\caption{Relative time-integrated $\elltwo$-error of the \methodAcronymROM.}
\label{fig:sod_error_methods_a}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/objective_vs_window_dtvar.pdf}
\caption{Time-integrated residual norm of the \methodAcronymROM.} 
\label{fig:sod_error_methods_b}
\end{subfigure}
\caption{Performance metrics of the Galerkin, LSPG, and \methodAcronymROMs.} 
\label{fig:time_step_study}
\end{center}
\end{figure}

@@

Figure~\ref{fig:convergence_study} shows the time-integrated $\ell^2$-error of the various ROM solutions, as well as that of the optimal solution as measured in the $\ell^2$-norm (i.e., the orthogonal projection of the FOM solution onto the trial subspace). In Figure~\ref{fig:sod_error_converge} we first observe that for a basis of dimension $\romdim = 10$, the Galerkin, LSPG, and WLS ($\Delta T = 0.002$, $\Delta T = 0.004$) ROMs do not converge. Upon investigation, this lack of convergence occurred due to the presence of negative pressure. We next observe that the time-integrated $\ell^2$-error monotonically decreases as the basis dimension increases for all remaining ROM solutions with the exception of the Galerkin, LSPG, and WLS ($\Delta T = 0.002$, $\Delta T = 0.004$) ROMs at the dimension $\romdim = 30$. Overall, we observe that larger window sizes yield lower errors when the basis dimension is small. For larger basis dimensions we observe all ROMs to yield accurate solutions, and a larger window size does not always yield a lower error. In Figure~\ref{fig:sod_residual_converge}, we next observe that the time-integrated residual norm monotonically decreases as the basis dimension grows. This result is supported, although not guaranteed, by the error analysis presented in Section~\ref{sec:analysis} (and Ref.~\cite{carlberg_lspg_v_galerkin} for LSPG). Finally, we observe that larger window sizes yield lower time-integrated residuals.

\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/error_converge.pdf}
\caption{Relative time-integrated $\elltwo$-error of the \methodAcronymROM.}
\label{fig:sod_error_converge}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/residual_converge.pdf}
\caption{Time-integrated residual norm of the \methodAcronymROM.} 
\label{fig:sod_residual_converge}
\end{subfigure}
\caption{Performance metrics of the Galerkin, LSPG, and \methodAcronymROMs\ as a function of basis dimension. Missing entries indicate that the solver did not converge.} 
\label{fig:convergence_study}
\end{center}
\end{figure}

Next, Figure~\ref{fig:convergence_study_pareto} presents Pareto plots for the time-integrated $\elltwo$-error and residual as a function of wall-clock time for ROMs with varying basis dimensions and window sizes. We observe intermediate window sizes (i.e., $\Delta T = 0.02,0.04$) to be Pareto-optimal for the error (i.e., lowest CPU time for a given error tolerance) for all runs with the exception of the $K=50$ case, in which case window sizes of $\Delta T = 0.002$ (i.e., LSPG) and $\Delta T = 0.004$ are Pareto-optimal. We next observe that intermediate window sizes are Pareto-optimal for the integrated residual norm for all cases, with the exception of runs using $\Delta T = 0.02, 0.04$, and $K=20$. Summing across the Pareto plots for the time-integrated error and residual, we find that --- for the window sizes and bases considered --- $\Delta T = 0.002$ (i.e., LSPG) is Pareto optimal three times, $\Delta T = 0.004$ one time, $\Delta T = 0.02$ 2 times, $\Delta T = 0.04$ 6 times, and $\Delta T = 0.2$ two times.

\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/pareto_error.pdf}
\caption{Relative time-integrated $\elltwo$-error of the \methodAcronymROM.}
\label{fig:sod_error_converge}
\end{subfigure}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/pareto_resids.pdf}
\caption{Time-integrated residual norm of the \methodAcronymROM.} 
\label{fig:sod_residual_converge}
\end{subfigure}
\caption{Pareto fronts for the time-integrate $\elltwo$-error (left) and time-integrated residual norm (right) as a function of wall-clock time.} 
\label{fig:convergence_study_pareto}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\begin{subfigure}[t]{0.45\textwidth}
\includegraphics[width=1.\linewidth]{figs/sod/converge_walltimes.pdf}
\caption{Relative time-integrated $\elltwo$-error of the \methodAcronymROM.}
\label{fig:sod_error_converge}
\end{subfigure}
\caption{Performance metrics of the Galerkin, LSPG, and \methodAcronymROMs\ as a function of basis dimension. Missing entries indicate that the solver did not converge.} 
\label{fig:convergence_study}
\end{center}
\end{figure}
